---
title: "DEBUNKING THE SIX FIGURE INCOME PROMISE"
author: "Cesar Y. Villarreal Guzman"
date: "October 19, 2020"
output: pdf_document
fontsize: 12pt
sansfont: Calibri Light
bibliography: biblio.bib
abstract: "Based on the recent social media advertisement trend of financial courses, we investigate the likelihood of having an annual income higher than or equal to $100,000CAD given young adult males who are self employed and work on average less than 30-40 hours a week. To accomplish this we fitted an elastic net logistic regression model using the Canadian 2016 General Social Survey and concluded on the unlikeliness of this event."
---

```{r, include=FALSE}
### Libraries ###
library(tidyverse)
library(caret)
library(glmnet)
library(cowplot)
# Load cleaned data
gss <- read.csv("gss.csv")


#### Generation of predictor variable plots
### Sex plot
sex_prop <- gss %>% group_by(sex) %>% summarise(n = n()) %>%
  mutate(prop = 100 * (n / sum(n))) 

p1 <- sex_prop %>%
  ggplot(aes(x="", y=prop, fill=sex)) +
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer(palette="Dark2") +
  theme_void() + 
  ylab("") + xlab("") + 
  labs(title = "", fill="Gender ") +
  geom_text(aes(label = paste0(round(prop), "%")), position = position_stack(vjust = 0.5), size = 4.7) +
  theme(text = element_text(size=15))

### Work hours plot
work_prop <- subset(gss, gss$nhours_pweek != "VS") %>% 
  group_by(nhours_pweek) %>% summarise(n = n()) %>%
  mutate(prop = 100 * (n / sum(n)))

p2 <- work_prop %>% ggplot(aes(x = "", y = prop, fill = nhours_pweek)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Accent") +
  theme_void() +
  ylab("") + xlab("") + 
  labs(title = "", fill="Work Hours") +
  geom_text(aes(label = paste0(round(prop), "%")), position = position_stack(vjust = 0.5), angle=45, size = 4.7) +
  theme(text = element_text(size=15))

### Age group plot
ageg_prop <- gss %>% group_by(resp_age_group) %>% summarise(n = n()) %>%
  mutate(prop = 100 * (n / sum(n)))

p3 <- ageg_prop %>% ggplot(aes(x = "", y = prop, fill = resp_age_group)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Accent") +
  theme_void() +
  ylab("") + xlab("") + 
  labs(title = "", fill="Age Groups") +
  geom_text(aes(label = paste0(round(prop), "%")), position = position_stack(vjust = 0.5), angle=45, size = 4.7) +
  theme(text = element_text(size=15))

### Age group plot
selfe_prop <- gss %>% group_by(self_empl) %>% summarise(n = n()) %>%
  mutate(prop = 100 * (n / sum(n)))

p4 <- selfe_prop %>% ggplot(aes(x = "", y = prop, fill = self_empl)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Greens") +
  theme_void() +
  ylab("") + xlab("") + 
  labs(title = "", fill="Self Employed") +
  geom_text(aes(label = paste0(round(prop), "%")), position = position_stack(vjust = 0.5), angle=45, size = 4.7) +
  theme(text = element_text(size=15))

### Income plot
income_prop <- gss %>% group_by(income) %>% summarise(n = n()) %>%
  mutate(prop = 100 * (n / sum(n)))

pincome <- income_prop %>% ggplot(aes(x = "", y = prop, fill = income)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "BrBG") +
  theme_void() +
  ylab("") + xlab("") + 
  labs(title = "", fill = "Income")+
  geom_text(aes(label = paste0(round(prop), "%")), position = position_stack(vjust = 0.5), angle=45) 
```


## Introduction

The year 2015 marked the beginning of the boom of the online education industry,
but more specifically financial education. The famous
advertisement video "Here In My Garage" by Tai Lopez, [^1] which at the moment has 
nearly 71 million views, displayed the immense potential of social media advertisement as
a viable marketing strategy. This particular video created a movement of the so called
"Financial Freedom" courses, which allegedly
offer a formula, a recipe, to not only making a six figure income, but
to do so by being self employed and working less time per week than the average.

[^1]: Link to video: https://www.youtube.com/watch?v=Cv1RJTHf5fk&ab_channel=TaiLopez

We would like to point out that this kind of "promises" are in fact not a new trend, 
there are many books on this topic which where published 
long before 2015. Some known authors include, 
Robert Kiyosaki and Tim Ferriss. 
In these past years we have seen new competitors
enter this "Financial Freedom" market via YouTube and other social media platforms,
we gave the example of Tai Lopez, but some other examples are Alex Becker and 
Dan Lok. 

In this investigation we answered the question: How likely are you
to make a six figure income in Canada? Moreover we discuss on the validity of these "Financial
Freedom" courses, since our results show that it is more likely that a young adult gets
an acceptance offer to Harvard University than to make a six figure income. Moreover
we are also interested in the number of worked hours per week, since most of these
courses claim that it is possible to have a high income while working 4-20 hours
per week. To be clear the range 4-20 is an average of many different advertisements on 
this topic.

Although there is no available evidence, one can make an educated guess based on 
YouTube's viewership demographics that most of these ads are being targeted to
young adult males, and hence the reason why we focus on this gender for our
discussion. However the model is fitted to work on any age group or gender.

To generate results that help answer our question, we look at the 
General Canadian Social Survey [@gss] the 2016 version, which 
had as focus "Canadians at Work and Home". We elaborate more on this data set
in the "Data" section. The main tool used for analysis in this investigation
is the elastic-net logistic regression model. 

## Data
The General Social Survey (GSS), as per the "Users Guide",
is a sample survey with cross-sectional design. 
The target population includes all persons 15 years of age and older in Canada,
excluding residents of the Yukon, Northwest Territories, and Nunavut and
full-time residents of institutions.[^2] The survey uses a frame that combines
telephone numbers with Statistics Canada’s Address Register. 

[^2]: Households without telephones were excluded from the survey population.

Sampling was done by dividing each of the ten provinces into strata with many
of the census metropolitan areas considered as separate strata. Each record in 
the survey frame was assigned to a stratum within its province. Then from each
stratum a selection of records was selected by simple random sampling. 
During collection, households that did not meet the eligibility criteria were 
terminated after an initial set of questions. If the household was eligible then
a respondent was randomly selected from each household to complete an electronic questionnaire
or to respond to a telephone interview.

The GSS aimed at collecting 20,000 samples but only achieved a sample size of $n=19,609$. 
The overall response rate was 50.8%, a very high response rate for a survey of 
this nature. Responses where adjusted using a weighting strategy which was followed
by standard bootstrap weighting.

The study, was divided into a first set of questions that had the purpose of 
demographic classification, questions like age, sex, marital status, relationship
of household members to respondent etc... The second section asked questions about
life at work, life at home, work life balance and health, well-being and resilience.
Income data was obtained through a linkage to tax data for respondents
who did not object to it. 

Data was retrieved from the Computing in the Humanities and Social Sciences (CHASS) 
at the University of Toronto website. A subset of the data was selected and retrieved
for this investigation. This subset included demographics, as well as income and work related
variables. The resulting data set contained 94 columns/features and
19609 observations. The data was cleaned by selecting wanted variables and 
remapping code values to text. Variables used for this report were: sex, age group, 
average number of work hours per week, income, and self employment. More information
on how to retrieve this data set and the code used to clean it can be found in the Appendix.
Figures 1 and 2 display the raw data used for the creation of the model.

```{r, echo=FALSE, warning=FALSE, fig.height = 12, fig.width = 11, error=FALSE, fig.align = "center", message=FALSE, fig.cap="GSS Distributions of Predictor Variables"}
# ggarrange(p1, p2, p3, p4, 
#           labels = c("Distribution of Sex", "Distribution of Income", 
#                      "Distribution of Age Groups", "Distribution of Self Employed Respondents"),
#           ncol =2, nrow = 2 )

ggdraw() +
  draw_plot(p1, x = 0, y = .5, width = .5, height = .5) +
  draw_plot(p4, x = .5, y = .5, width = .5, height = 0.5) +
  draw_plot(p2, x = 0, y = 0, width = .5, height = 0.5) + 
  draw_plot(p3, x= .5, y = 0, width = .5, height = 0.5) + 
  draw_plot_label(label = c("Distribution of Gender", "Distribution of Weekly Worked Hours",
                     "Distribution of Age Groups", "Distribution of Self Employed Respondents"), size = 15,
                  x = c(0, 0, 0.5, 0.3), y = c(1, 0.5, 0.5, 1))
```
\pagebreak
```{r, echo=FALSE, warning=FALSE, fig.height = 4, fig.width = 5, error=FALSE, fig.align = "center", message=FALSE, fig.cap="GSS Distributions of Income."}
pincome
```

## Model

In the introduction we outlined that we seek to model the likelihood of someone
having a six figure income based on sex, age, number of hours worked per week
and, if this individual is self employed. Fortunately, the GSS data set contains
all of these variables except for age. To solve this issue we use the age group
of respondents. Moreover, a new vector $\mathbf{y}=(y_1,\dots, y_n)$ was generated
for which $y_i=1$ represents that respondent $i$ has an income higher than or equal to $100,000CAD and 
$y_i=0$ otherwise. From this point on we will denote 
$\mathbf{X}=(\mathbf{x}_1,\dots, \mathbf{x}_n)$ 
as the matrix were observation $\mathbf{x}_j=(x_{j1}, \dots, x_{jp})^T$ has
$p$ predictor features. In this particular case we have 4 predictor features.

The original question is then simplified to the problem of predicting the binary
value $y$ given $\mathbf{x}$. An implicit assumption of a linear relationship is made here
to go along the research question and the opening discussion of the paper.
Therefore, our model of choice, which is perhaps the most used and
suited for binary classification, is the logistic regression model. 

Logistic regression arises from the desire to model the posterior probability of the two classes $0$ 
and $1$ via linear functions. Suppose that the probability of having a
six figure income is equivalent to $p$, then the model has the
form
\[\log\frac{p}{1-p}=\beta_0+\beta^T\mathbf{x}\]
here we assume $\beta=(\beta_1, \dots, \beta_4)^T$ is a vector of appropriate size for left multiplication
with $\mathbf{x}$. Logistic regression models are usually fit by maximum likelihood.[@hastie_friedman_tisbshirani]

In particular we use a regularized version of the logistic regression model. This is 
because, generally speaking regularization tends to be beneficial for the predictive 
performance of a model. There are three common methods for regression regularization:
lasso [@tibshirani], ridge [@hoerl_kennard] and elastic net regression [@zou_hastie]. 

Given strong prior knowledge it may be better to pick lasso or ridge in 
place of elastic net. However, in absence of prior knowledge, 
elastic net should be the preferred solution and our regularization method
for this problem. The limitation here is that elastic net has significantly higher
computational complexity and is therefore much slower than lasso or ridge. Both
ridge and lasso represent viable alternatives to elastic net regularization and
for the model as a whole.

Elastic net regression combines both strengths and solves the limitations of lasso and ridge regression. 
This is achieved by combining the $L_1$ and $L_2$ penalties. Following
the notation used above, the log-likelihood for the conventional logistic regression model
for $N$ examples is given by
\[\ell(\beta_0, \beta)=\sum_{i=1}^Ny_i\log p + (1-y_i)\log(1-p)\]
\[=\sum_{i=1}^Ny_i(\beta_0+\beta^T\mathbf{x}_i)-\log(1 + exp\{\beta_0+\beta^T\mathbf{x}_i\})\]
then, the elastic net logistic regression model has the following
log-likelihood for $N$ examples
\[\ell_{elnet}(\beta_0, \beta)=\left[\sum_{i=1}^Ny_i(\beta_0+\beta^T\mathbf{x}_i)-\log(1 + exp\{\beta_0+\beta^T\mathbf{x}_i\})\right] - \left[\lambda_1\sum_{j=1}^4|\beta_j| + \lambda\sum_{j=1}^4\beta_j^2\right]\]
in the equation above we assume that $\beta=(\beta_1,\dots, \beta_4)$. Also in this
new equation, the value $\lambda_1\sum_{j=1}^4|\beta_j|$ is the $L_1$ regularization penalty, and
similarly $\lambda\sum_{j=1}^4\beta_j^2$ the $L_2$ regularization penalty.
We use this new function to compute the pair $(\hat{\beta_0}, \hat{\beta})$ given by
\[(\hat{\beta_0}, \hat{\beta})=\text{arg}\max_{(\beta_0, \beta)} \ell_{elnet}(\beta_0, \beta)\]
Or equivalently
\[(\hat{\beta_0}, \hat{\beta})=\text{arg}\min_{(\beta_0, \beta)} -\ell_{elnet}(\beta_0, \beta)\]
Typically algorithms that solve this optimization problem tend to converge since
log-likelihood is concave.

We use the implementation of this model found in the R package **glmnet** [@glmnet_cit].
For the training phase we used a 80-20 train to test ratio split for 
the GSS data set. That is 80% of the data was used to fit the model and the
remaining 20% was used to evaluate the performance of the model. Since 
all variables are categorical, we required the creation of several dummy binary
variables, also known as "one-hot" encoding.

```{r, include = FALSE}
gss2 <- gss[, c("resp_age_group", "sex", "nhours_pweek", "self_empl", "income")]

set.seed(12345)
# Create a 80-20 train-test sample split
train.sample <- gss2$income %>%
  createDataPartition(p = 0.8, list = FALSE)

# Create binary variable Y and remove income column
gss2 <- gss2 %>%  mutate(income100 = ifelse(income == "$100k-$124,999" | income == "$125k+", 1, 0))
gss2 <- gss2[, -5]

# Split data set
gss.train <- gss2[train.sample, ]
gss.test <- gss2[-train.sample, ]

# Define X and Y for training
x <- model.matrix(income100~., gss.train)[,-1]
y <- gss.train$income100

# Find best lambda using cross-validation
cv_elnet <- cv.glmnet(x, y, family = "binomial", alpha = 0.3)

# Fit the final model on the training data
gss_model <- glmnet(x, y, alpha=0.3, family="binomial", lambda = cv_elnet$lambda.min)

# Test model on the test set
x.test <- model.matrix(income100~., gss.test)[,-1]
prob_predictions <- gss_model %>% predict(newx = x.test, type="response")

# We pick a small threshold based on the proportions of people with high incomes
predictions <- ifelse(prob_predictions > 0.1, 1, 0)

# Select vector of responses for model evaluation
y.test <- gss.test$income100

# The function confusionMatrix creates a list of classification metrics
# among them, a confusion matrix. Some other metrics include accuracy, precision
# recall etc...
accuracy <- mean(y.test == predictions)
tmp_coeffs <- coef(gss_model)

lvs <- c(1, 0)
true <- factor(y.test, levels = lvs)
pred <- factor(predictions, levels = lvs)

# From the caret package create metrics for model evaluation
metrics <- confusionMatrix(pred, true)
```


## Results
```{r, echo=FALSE, }
data.frame(name = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1], coefficient = tmp_coeffs@x) %>%
  knitr::kable(digits = 2, col.names = c("Variable Name", "Coefficient"),
               caption = "Coefficient Values for Fitted GSS Model")
```

Table 1 displays the fitted model for our classification problem. It is important
to note that Table 1 does not include any standard errors of regression 
coefficients or other estimated quantities. This is because as @glmnet_cit 
explain in their implementation notes "...this package deliberately does not provide them. The reason for this is that standard errors are not very meaningful for strongly biased estimates such as arise from penalized estimation methods."

The accuracy on the test set was of **`r round(accuracy * 100, digits=2)`%**,
with sensitivity **`r round(100*metrics$byClass[["Sensitivity"]], digits=2)`%** and specificity
**`r round(100* metrics$byClass[["Specificity"]], digits = 2)`%**. Sensitivity, in this problem,
measures the proportion of high income individuals that are correctly identified, and
specificity measures the proportion of non-high income individuals that are correctly identified.
We also include a confusion matrix plot, Figure 3, of our model on the test set which can be found in the 
"Discussion" section.


```{r, echo=FALSE, warning=FALSE, fig.height = 4, fig.width = 5, error=FALSE, fig.align = "center", message=FALSE, fig.cap="Confusion Matrix of GSS Model on Test Set."}
# Select confusion matrix
table <- data.frame(metrics$table)

plotTable <- table %>%
  mutate(goodbad = ifelse(table$Prediction == table$Reference, "good", "bad")) %>%
  group_by(Reference) %>%
  mutate(prop = Freq/sum(Freq))

# fill alpha relative to sensitivity/specificity by proportional outcomes within reference groups (see dplyr code above as well as original confusion matrix for comparison)
ggplot(data = plotTable, mapping = aes(x = Reference, y = Prediction, fill = goodbad, alpha = prop)) +
  geom_tile() +
  geom_text(aes(label = Freq), vjust = .5, fontface  = "bold", alpha = 1) +
  scale_fill_manual(values = c(good = "green", bad = "red")) +
  theme_minimal() +
  xlim(rev(levels(table$Reference))) + 
  xlab("True Values") + labs(fill = "GoodBad", alpha = "Proportion") +
  theme(legend.position="none")
```

## Discussion

In statistical terms, Figure 3 serves as evidence that our model lacked precision
because of a high number of false positives.
However, our achieved model accuracy is much better than chance which is closer to 50%.

With regards to the coefficients, it is no surprise that the most significant values are
for the older age groups. It is reasonable to assume that as one gets older the 
chances of having a higher position at a company, or simply having profitable 
investments is higher. It is also no surprise that there is a positive relationship between number of hours 
and income. In a country like Canada where jobs are usually compensated based on 
hourly work this is what's to be expected.

What's interesting, however, is the little effect of the self employment
variable on this model. The fact that the self employed group are 
a minority in most societies displays itself in this result.

Some of the weaknesses of our approach come from the limitations of using linear models 
on an apparent non-linear problem.
As an alternative, we suggest fitting a support vector machine model.
This method has the benefit of being partially linear but breaking 
its linearity with what is referred to as the "kernel trick". Another 
alternative is to fit a decision tree classifier.

Similarly, a limitation of our approach is that we do not posses any data on 
entrepreneurs who fit the demographic profiles. Some areas of future work
are to sample a population of entrepreneurs and ask questions regarding the 
amount of time it took to start generating a six figure income, assuming they
already make six figures. We focused on young adult males leaving behind
other genders. Another area of future work could be to apply the same research 
question to females and compare the results. 

Finally, as per the introduction, consider
a young adult male in the age group 25-34 who is self 
employed and who works less than the average of 30-40 hours. Our model then estimates 
that the probability of having a six figure income is of
**`r round(prob_predictions[6] * 100, digits=2)`%**, a very unlikely outcome. 
On the contrary we can say that our model shows evidence that the probability of having a high 
income is increased given an older person, working more than the average and not being self employed.


## Appendix
Data was retrieved from the Computing in the Humanities and Social Sciences (CHASS) 
at the University of Toronto website. In order to retrieve this data set from the 
CHASS website, a valid University of Toronto email is required. Other academic 
institutions offer this data set but instructions will not be included.

The official link to the CHASS website is **http://dc.chass.utoronto.ca/myaccess.html**, on
arrival to the site click on the option, located on the side bar, "SDA @ CHASS". 
Login with your University of Toronto credentials. Select your language of choice 
and on the keyboard click "Ctrl + f", Command for Mac users, and type "gss". Hit 
enter and click on "General Social Survey". Locate the 2016 version and click on "data", then
on the top bar, under "Download" click "Customized Subset". We recommend downloading
a "csv" file together with the codebook. Our cleaning code does not require any
data definitions file. To ensure our code works we advise to select "All" 
variables. Variables of interest can be selected instead of "All" however our 
code may not work. At the bottom click continue and download all files.

All cleaning and model code used for this project can be found here:


**https://github.com/cesar-yoab/elnet_regression_R**

---
nocite: '@*'
---

\pagebreak
\pagebreak

## References